HO unification from object language to meta language

### 2024-03-25 paper/appendix.tex
more on 5

  type solve-links links -> links -> ho.subst -> ho.subst -> o.
  solve-links [] [] X X.
  solve-links [A|L1] [A|L3] S S2 :- take-link A L1 B L2, !,
    same-link-eta A B S S1, 
    solve-links L2 L3 S1 S2.
  solve-links [L0|L1] L3 S S2 :- deref-link S L0 L,
    solve-link-abs L R S S1, !,
    solve-links L1 L2 S1 S2, append R L2 L3.
\end{elpicode}

>>>>

  type progress1 links -> links -> ho.subst -> ho.subst -> o.
  progress1 [] [] X X.
  progress1 [A|L1] [A|L3] S S2 :- take-link A L1 B L2, !,
    same-link-eta A B S S1, 
    progress1 L2 L3 S1 S2.
  progress1 [L0|L1] L3 S S2 :- deref-link S L0 L,
    solve-link-abs L R S S1, !,
    progress1 L1 L2 S1 S2, append R L2 L3.
\end{elpicode}

### 2024-03-25 paper/main.tex
more on 5

\begin{proposition}[Simulation]\label{prop:sumul}
$\forall \mathcal{S}, \forall \mathcal{N}$
$$
  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[Properties of \Uo{}]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
\unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}\\
% \forall \rho'\rho,
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \llambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2 \label{prop:simulation}
\end{gather}
\end{proposition}

@@

Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

@@

@@

\input{code/comp_base2}

\noindent
The auxiliary function \elpiIn{close-links} tests if the bound variable
\elpiIn{v} really occurs in the link. If it is the case the link is wrapped into
an additional \elpiIn{abs} node binding \elpiIn{v}. In this way links generated
deep inside the compiled terms can be moved outside their original context
of binders.

Note that \elpiIn{pi x y\..} is syntactic sugar for \elpiIn{pi x\ pi y\..}.

@@

XXX links are update unlike section 2
\input{code/hstep}

note \elpiIn{(a = b) c d} is \elpiIn{(=) a b c d}.

@@

\subsection{Example}

OK

\begin{elpicode}
Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
      , flam x\ fapp[fcon"g", fcon"a"] ]
Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
link z z (s z)
HS = [some (abs x\con"a")]
S = [some (flam x\fcon a)]
\end{elpicode}

KO

\begin{elpicode}
  Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
  , flam x\ fapp[fcon"g", fcon"a"] ]
Problems = [ pr 0 1   % $A = \lambda x.x$
           , pr 2 3 ] % $A a = a$
lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
link z z (s z)
HS = [some (abs x\con"a")]
S = [some (flam x\fcon a)]
lam x\ app[f, app[X, x]] = Y,
  lam x\ x) = X.
\end{elpicode}

\otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
\otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
\otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
\otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
\otext{Links used to instantiate vars of elpi}
\otext{After all links, the solution in links are compacted and given to coq}
\otext{It is not so simple, see next sections (multi-vars, eta, beta)}

The compilation step is meant to recover the higher-order variables of the OL,
expressed in a first order way, by replacing them with higher-order variables in
the ML. In particular, every time a variable of the OL is encountered in the
original term, it is replaced with a meta variable, and if the OL variable is
applied to a list of distinct names $L$, then this list becomes the scope of the variable.
For all the other constructors of
\elpiIn{tm}, the same term constructor is returned and its arguments are
recursively compiled. The predicate in charge for term compilation is:

\elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.

\noindent
where, we take the term of the OL, produce the term of the ML, take a list
of link and produce a list of new links, take a substitution and return a
new substitution.

In particular, due to programming constraints, we need to drag the old subst and
return a new one extended, if needed, with the new declared meta-variables.

The following code
%
\begin{elpicode}
  kind link type.
  type link nat -> nat -> nat -> subst.
\end{elpicode}
%
\noindent
defines a link, which is a relation between to variables indexes, the first
being the index of a OL variable and the second being the index of a ML
variable. The third integer\todo{integer or nat?} is the number of term in the
scope of the two variables, or equivalently, in a typed language, their arity.

As an example, let's study the following unification problem (a slightly
modified version from \cref{sec:intro}):

\begin{elpicode}
  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
    lam x\ app [c"decision", app[uv 0, x]]
\end{elpicode}

\noindent
we have the main unification problem where the nested \elpiIn{app} nodes have
lists of different lengths making the unification to fail. The compilation of
these terms produces a new unification problem with the following shape:

\begin{elpicode}
  lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~
    lam x\ app [c"decision", uv 1 [x]]
\end{elpicode}

\noindent
The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
ML, that is, the index \elpiIn{0} for that unification variable of the OL term
has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
different substitution mapping, one for the OL and one for the ML and the indexes
of variable point to the respective substitution.

decomp che mappa abs verso lam
\noindent
\otext{An other example: \\
  \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}

>>>>

\begin{proposition}[Simulation]\label{prop:simulation}
$\forall \mathcal{S}, \forall \mathcal{N},$
$$
  \runF{\mathcal{S}}{\mathcal{N}}{\rho}
  \Leftrightarrow
  \runH{\mathcal{S}}{\mathcal{N}}{\rho}
$$
\end{proposition}

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N},$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[Properties of \Uo{}]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\customlabel{prop:correct}{(correct)}\\
\unifcomplete{\Eo}{\Uo}{s}\customlabel{prop:complete}{(complete)}\\
% \forall \rho'\rho,
  \rho s_1 \Eo \rho s_2 \Rightarrow
  \rho' \subseteq \rho \Rightarrow
  \rho's_i \in \llambda \Rightarrow
  \rho' s_1 \Uo \rho' s_2 %\label{prop:simulation}
\end{gather}
\end{proposition}

@@

\marginpar{fix}Property \ref{prop:simulation} states that \Uo, hence our compilation scheme,
is resilient to unification problems outside \llambda solved by
a third party. We believe this property is of practical interest since we
want the user to be able to add heuristics via hand written rules
to the ones obtained by our compilation scheme. A Typical example
is the following problem \ref{problem:q} that is outside \llambda:

@@

\noindent
When it is clear from the context we shall use the same syntax for \Fo{} terms
(although we never subscripts unification variables).

@@

\subsection{Compilation}
\marginpar{manca beta normal in entrata}

@@

\input{code/comp_lam}

\noindent
In the code above the syntax \elpiIn{pi x y\..} is syntactic sugar for
iterated \elpiIn{pi} abstraction, as in \elpiIn{pi x\ pi y\..}.

The auxiliary function \elpiIn{close-links} tests if the bound variable
\elpiIn{v} really occurs in the link. If it is the case the link is wrapped into
an additional \elpiIn{abs} node binding \elpiIn{v}. In this way links generated
deep inside the compiled terms can be moved outside their original context
of binders.

\input{code/comp_close_links}

\noindent
Note that we could remove the second rule, whose purpose is to make
links more readable by pruning unneeded abstractions (unused context entries).

@@

XXX links are update unlike section 2
\input{code/hstep}
\input{code/progress}

\noindent
Note thar \elpiIn{((A ~\Ue~B) C D)} is syntactic sugar for
\elpiIn{((~\Ue~) A B C D)}.

@@

\subsection{Definition of \Uo and its properties}

\input{code/unif_fo}

The code given so far applies to terms in $\beta\eta$-normal form where
unification variables in \Fo{} can occur non linearly but always with
the same number of arguments, and where their arguments are distinct names
(as per \llambda).

\begin{lemma}[Compilation round trip] If
  \elpiIn{comp S T [] M [] _ [] _} then \elpiIn{decomp M T S}
\end{lemma}
\begin{proof}[Proof sketch]
trivial, since the terms are beta normal beta just builds an app.
\end{proof}

\begin{lemma}
Properties \ref{prop:correct} and
\ref{prop:complete} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo on ground terms. What we have to show is that whenever two different \Fo
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can find this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the \Fo{} terms, the are two interesting cases:
\begin{itemize}
\item \elpiIn{fuva X ~\Uo~ s}. In this case after \elpiIn{comp} we have
  $Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
  $\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$.
\item \elpiIn{fall[fuva X|L] ~\Uo~ s}. In this case
 we have $Y_{\vec{x}} \Ue t$ that succeeds with
 $\sigma = \{ \vec{y} \vdash Y \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
 is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
 Thanks to \ref{clause:beta1}
 $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
\end{itemize}
Since the mapping is a bijection occur check in \Ho{} corresponds to occur
check in \Fo{}.
\end{proof}

\begin{lemma} Properties simulation (\ref{prop:simulation}) and
fidelity (\ref{prop:fidelity}) hold
\end{lemma}
\begin{proof}[Proof sketch]
Since \elpiIn{progress1} is trivial \fstep and \hstep are the same, that is
in this
context where input terms are $\beta\eta$-normal and we disregard $\eta$-equivalence
\Ue is equivalent to \Uo.
\end{proof}

\subsection{Limitations of by this basic scheme}

\begin{gather}
\lambda x y.F~y~x = \lambda x y.x\\
\lambda x. f~(F~x)~x = f~(\lambda y.y)
\end{gather}
Note that here $F$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover that
$F = \lambda x\lambda y.y$ (i.e. that $F$ discards the $x$ argument).
Both problems are addressed in the next section.

% \subsection{Example}

% OK

% \begin{elpicode}
% Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
%       , flam x\ fapp[fcon"g", fcon"a"] ]
% Problems = [ pr z (s z) ] % $\lambda x.g (F x) = \lambda x.g a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (flam x\fcon a)]
% \end{elpicode}

% KO

% \begin{elpicode}
%   Terms [ flam x\ fapp[fcon"g",fapp[fuva z, x]]
%   , flam x\ fapp[fcon"g", fcon"a"] ]
% Problems = [ pr 0 1   % $A = \lambda x.x$
%            , pr 2 3 ] % $A a = a$
% lam x\ app[con"g",uva z [x]] ~\Uo~lam x\ app[con"g", con"a"]
% link z z (s z)
% HS = [some (abs x\con"a")]
% S = [some (flam x\fcon a)]
% lam x\ app[f, app[X, x]] = Y,
%   lam x\ x) = X.
% \end{elpicode}

% \otext{Goal: $s_1 \Uo s_2$ is compiled into $t_1 \Ue t_2$}
% \otext{What is done: uvars \elpiIn{fo_uv} of OL are replaced into uvars \elpiIn{ho_uv} of the ML}
% \otext{Each \elpiIn{fo_uv} is linked to an \elpiIn{ho_uv} of the OL}
% \otext{Example needing the compiler v0 (tra l'altro lo scope è ignorato):\\ \elpiIn{lam x\ app[con"g",app[uv 0, x]] ~\Uo~lam x\ app[con"g", c"a"]}}
% \otext{Links used to instantiate vars of elpi}
% \otext{After all links, the solution in links are compacted and given to coq}
% \otext{It is not so simple, see next sections (multi-vars, eta, beta)}

% The compilation step is meant to recover the higher-order variables of the OL,
% expressed in a first order way, by replacing them with higher-order variables in
% the ML. In particular, every time a variable of the OL is encountered in the
% original term, it is replaced with a meta variable, and if the OL variable is
% applied to a list of distinct names $L$, then this list becomes the scope of the variable.
% For all the other constructors of
% \elpiIn{tm}, the same term constructor is returned and its arguments are
% recursively compiled. The predicate in charge for term compilation is:

% \elpiIn{type comp tm -> tm -> links -> links -> subst -> subst -> o}.

% \noindent
% where, we take the term of the OL, produce the term of the ML, take a list
% of link and produce a list of new links, take a substitution and return a
% new substitution.

% In particular, due to programming constraints, we need to drag the old subst and
% return a new one extended, if needed, with the new declared meta-variables.

% The following code
% %
% \begin{elpicode}
%   kind link type.
%   type link nat -> nat -> nat -> subst.
% \end{elpicode}
% %
% \noindent
% defines a link, which is a relation between to variables indexes, the first
% being the index of a OL variable and the second being the index of a ML
% variable. The third integer\todo{integer or nat?} is the number of term in the
% scope of the two variables, or equivalently, in a typed language, their arity.

% As an example, let's study the following unification problem (a slightly
% modified version from \cref{sec:intro}):

% \begin{elpicode}
%   lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Uo~
%     lam x\ app [c"decision", app[uv 0, x]]
% \end{elpicode}

% \noindent
% we have the main unification problem where the nested \elpiIn{app} nodes have
% lists of different lengths making the unification to fail. The compilation of
% these terms produces a new unification problem with the following shape:

% \begin{elpicode}
%   lam x\ app[c"decision", app[c"nfact", x, c"3"]] ~\Ue~
%     lam x\ app [c"decision", uv 1 [x]]
% \end{elpicode}

% \noindent
% The main difference is the replacement of the subterm \elpiIn{app[uv 0, x]} of
% the OL with the subterm \elpiIn{uv 0 [x]}. Variable indexes are chosen by the
% ML, that is, the index \elpiIn{0} for that unification variable of the OL term
% has not the sam meaning of the index \elpiIn{0} in the ML. There exists two
% different substitution mapping, one for the OL and one for the ML and the indexes
% of variable point to the respective substitution.

% decomp che mappa abs verso lam
% \noindent
% \otext{An other example: \\
%   \elpiIn{lam x\ app[f, app[X, x]] = Y, (lam x\ x) = X.}}
